---
title: "Data analysis"
author: "Luis Fernando Delgado"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries

```{r}
rm(list = ls())
library(asreml)
source("https://raw.githubusercontent.com/Cassava2050/PPD/main/utilities_tidy.R")

trial_interest <- "CQEPR"
year_interest <- 2023
```

## master_data to save the results

```{r}
master_data <- list()
```

## Load the tidy data

```{r}
trial_set_number = 1
# all files in the folder
list_file = list.files(here::here("output"))
# tidy data of the trials interested
sel_file = list_file[str_detect(list_file, "_tidy_data4analysis_") &
                       str_detect(list_file,
                                  paste(year_interest, trial_interest, sep=""))]

sel_file
# the data we will use
sel_file_use = sel_file[1]

sel_file_use
trial1_tidy = read.csv(here::here("output", sel_file_use), header=TRUE,
                        stringsAsFactors = FALSE,
                        as.is=T,
                        check.names = FALSE)
if(trial_set_number == 1){
  trial_tidy_all = trial1_tidy
}
```

## Filter CQEPR trials

```{r}
trial_tidy_all <- trial_tidy_all %>% filter(str_detect(trial_name, trial_interest))
```


## Obtain all the trait information using a cloud file (gitHub) -------

```{r}
trait_all <-
  read.csv("https://raw.githubusercontent.com/lfdelgadom/standar_col_names_CB/main/standar_col_names.csv") %>%
  select(analysis_col_name) %>%
  filter(str_detect(analysis_col_name, "obs_"))
trait_all_adj <- gsub("obs_", "", trait_all$analysis_col_name)
trait_all_adj = c(trait_all_adj,
                  "harvest_number_plan", "germination_perc",
                  "yield_ha_v2", "DM_yield_ha", "starch_content")
trait_all_adj <- gsub("-", "_", trait_all_adj)
```

## Meta info.

```{r}
meta_all <-
  read.csv("https://raw.githubusercontent.com/lfdelgadom/standar_col_names_CB/main/standar_col_names.csv") %>%
  select(analysis_col_name) %>%
  filter(str_detect(analysis_col_name, "use_"))
meta_all_adj <- gsub("use_", "", meta_all$analysis_col_name)
meta_all_adj <- c(
  meta_all_adj,
  "check_released", "latitude", "longitude",
  "altitude", "department", "country",
  "ag_zone", "location_short"
)
```

## Select the observations for analysis

```{r}
names(trial_tidy_all) <- gsub("-", "_", names(trial_tidy_all))
analysis_trait <- names(trial_tidy_all)[names(trial_tidy_all) %in% trait_all_adj]
print("All the traits investigated:")
print(analysis_trait)

```

## Select the meta information for analysis

```{r}
meta_col <- names(trial_tidy_all)[names(trial_tidy_all) %in% meta_all_adj]
print("All the meta information:")
print(meta_col)
```

## Check the SD of each trait

```{r}
trial_rm_sd <- remove_no_var_tidy(my_dat = trial_tidy_all,
                   analysis_trait = analysis_trait,
                   meta_info = meta_col)
master_data[["mean_of_sd"]] = sd_mean
```

## Trait ideal

```{r}
no_traits_for_analysis <- c("stake_plant" , "planted_number_plot", 
                            "harvest_number", "root_weight_air", 
                            "root_weight_water", "harvest_number_plan",
                            "yield_ha_v2", "root_rot_perc", "germinated_number_plot",
                            "frogskin1_5"
)
trait_ideal <- analysis_trait[!analysis_trait %in% no_traits_for_analysis]
print("the trait ideal is:"); trait_ideal

trait_ideal %>% as.data.frame() %>% write.table("clipboard", sep = "\t", row.names = F)
```

# Genotypic correlation (Phenotypic values)

```{r}
pheno_cor <- trial_rm_sd %>% select(all_of(trait_ideal))
correlation <- gg_cor(
  colours = c("red", "white", "blue"),
  data = pheno_cor,
  label_size = 2
  )
ggsave(paste("images\\pheno_corr", trial_interest, ".png", sep = "_"),
     plot = correlation, units = "in", dpi = 300, width = 12, height = 8
   )
```


## Check design experimental

### Agriutilities library

```{r}
my_dat <- trial_rm_sd %>% 
  add_column(block = NA) %>% mutate(block = as.factor(block))

results <- check_design_met(
  data = my_dat,
  genotype = "accession_name",
  trial = "trial_name",
  traits = trait_ideal,
  rep = "rep_number",
  col = "col_number",
  row = "row_number",
  block = "block"
)

shared <- plot(results, type = "connectivity")

ggsave(paste('images\\shared_', trial_interest, ".png", sep = "_"),
       plot = shared, units = "in", dpi = 300, width = 8, height = 6)
  
summary <- results$summ_traits 

p1 <- summary %>% 
  ggplot(aes(x = traits , y = trial_name, label = round(miss_perc,2),  fill = miss_perc ))+
  geom_tile(color = "gray")+
  geom_text(color = "white")+
  theme_minimal(base_size = 13)+
  labs(title = "Percentage of missing values (exp/trait)", x = "", y = "") +
  theme(axis.text.x = element_text(hjust = 1 , angle = 75, size = 16),
        axis.text.y = element_text(size = 16))
p1
ggsave(paste0("images\\missing_", trial_interest, "_", ".png"),
     plot = p1, units = "in", dpi = 300, width = 15, height = 6
   )
master_data[["summ_traits"]] <- summary
```

## Single trial analysis

I should test removing the trials with non check clones. 

```{r}
obj <- single_trial_analysis(results = results,
                             progress = TRUE,
                             remove_outliers = FALSE)

trials <- unique(my_dat$trial_name)

header_sort = vector()
i = 1
for (i in 1:length(trials)) {
  
  cat("\n_______________")
  cat("\nTRIAL:", trials[i], "\n")
  cat("_______________\n")
  
  for (j in 1:length(trait_ideal)) {
    
    blue_blup <- obj$blues_blups %>% 
      filter(trial == trials[i]) %>% 
      select(-c(trial, seBLUEs, seBLUPs, wt)) %>% 
      pivot_wider(names_from = "trait", values_from = c("BLUEs", "BLUPs"))
    
    header_sort = c(header_sort,
                  grep(trait_ideal[j], sort(names(blue_blup)), value=TRUE))
  blue_blup <- blue_blup %>% dplyr::select(genotype, any_of(header_sort)) %>% 
    mutate(across(where(is.double), round, 1))
  }
 master_data[[paste0("BLUP_BLUE_", trials[i])]] <- blue_blup
}


# save into pdf spatial plots
if(TRUE){
  pdf(paste(here::here("output//"), "01_", trial_interest, "_spatial_plot_", 
            ".pdf", sep = ""), width = 8, height = 6)
  
  plot(obj, type = "spatial") 
  
  dev.off()
}

master_data$BLUP_BLUE_202339CQEPR_cere
```



## Single heritability

```{r}
single_h2 <- obj$resum_fitted_model[ ,1:3] %>% 
  group_by(trial) %>%
  spread(trait, value = heritability) 

master_data[["single_h2"]] <- single_h2 

single_h2 %>% 
  write.table("clipboard", sep = "\t", col.names = T, row.names = F, na = "")


# single heritability plot
H2 <- obj$resum_fitted_model[ ,1:3] %>% 
  ggplot(aes(x = trait, y = trial, label = heritability, fill = heritability)) +
  geom_tile(color = "black") +  # This will also avoid drawing borders for NA tiles
  geom_text(color = "white") +  # Avoids placing text on NA tiles
  labs(title = "Broad sense heritability", x = "", y = "") +
  theme_xiaofei() +
  scale_x_discrete(position = "top") +
  theme(axis.text.x = element_text(angle = 75, hjust = 0, size = 10),
        axis.text.y = element_text(size = 13)) 
H2
ggsave(paste0("images\\H2", trial_interest, "_", ".png"),
     plot = H2, units = "in", dpi = 300, width = 15, height = 6
   )
  
```

# Germination plot 

```{r}
# 202330BCEPR_stom
my_dat %>% filter(trial_name == "202339CQEPR_cere") %>% 
  ggplot(aes(x = reorder(accession_name, -germination_perc), 
             y = germination_perc, fill = factor(rep_number))) +
  facet_grid(factor(rep_number)~trial_name) +
  geom_col(position = "dodge") +
  theme_xiaofei() +
  theme(axis.text.x = element_text(size = 5)) +
  labs(x = "accession_name", fill = "rep_number")

ggsave(paste0("images\\germ_cere", trial_interest, "_", ".png"),
        units = "in", dpi = 300, width = 15, height = 6
   )


# 202338CQEPR_repe
my_dat %>% filter(trial_name == "202338CQEPR_repe") %>% 
  ggplot(aes(x = reorder(accession_name, -germination_perc), 
             y = germination_perc, fill = factor(rep_number))) +
  facet_grid(factor(rep_number)~trial_name) +
  geom_col(position = "dodge") +
  theme_xiaofei() +
    theme(axis.text.x = element_text(size = 5)) +
  labs(x = "accession_name", fill = "rep_number")

ggsave(paste0("images\\germ_repe", trial_interest, "_", ".png"),
       units = "in", dpi = 300, width = 15, height = 6
   )
```

## Single CV

```{r}
single_CV <- obj$resum_fitted_model[ ,c(1:2, 4)] %>% 
  group_by(trial) %>%
  spread(trait, value = CV) 

master_data[["single_CV"]] <- single_CV 
```

## Single variance components

```{r}
single_var_comp <- obj$resum_fitted_model[ ,c(1:2, 5:6)]

master_data[["single_var_comp"]] <- single_var_comp 
```


## Multi environmetal analysis

```{r}
# several trials
# met_results <- met_analysis(obj,
#   filter_traits =
#     trait_ideal[!trait_ideal %in% c(
#       "CBB_3mon",
#       "branch_number",
#       "height_1st_branch",
#       "height",
#       "height_wt_leaf",
#       "lodging1_3_6mon", "root_constriction1_3",
#       "root_length1_3",
#       "root_number_commercial",
#       "root_peduncle1_3",
#       "root_shape1_6",
#       "root_skin_color1_3",
#       "vigor1_5",
#       "lodging1_3",
#       "root_rot_number"
#     )],
#   h2_filter = 0.1,
#   progress = TRUE,
#   # vcov = "fa2"
# )

met_results <- met_analysis(obj,
  filter_traits = "germination_perc",
  h2_filter = 0.1,
  progress = TRUE,
  # vcov = "fa2"
)


# h2 gxe
master_data[["h2_gxe"]] <- 
  met_results$heritability %>% 
  arrange(desc(h2)) %>%
  mutate(across(where(is.numeric), round, 2))

master_data$h2_gxe %>%
  write.table("clipboard", col.names = T, row.names = F, sep = "\t")

# BLUPs gxe
BLUPs_table <- 
  met_results$overall_BLUPs %>% 
  select(-c(std.error, status)) %>% 
  group_by(genotype) %>% 
  spread(trait, value = predicted.value) %>% 
  rename("accession_name" = genotype) %>% 
  mutate(across(where(is.numeric), round, 2)) %>% 
  ungroup() 
#save the BLUPs data
master_data[[paste0("BLUPs_", "gxe")]] <- BLUPs_table
```

## Genotypic Correlation: Locations

```{r}
# Define a function to create plots and save them
plot_and_save <- function(trait, file_name, met_results, trial_interest) {
  covcor_heat(matrix = met_results$VCOV[[trait]]$CORR, size = 4, legend = c(0.35, 0.8)) +
    theme(axis.text.y = element_text(size = 14),
          axis.text.x = element_text(size = 14))
  
  # Construct file path using the file_name and trial_interest
  file_path <- paste0("images\\", file_name, "_vcor", trial_interest, Sys.Date(), ".png")
  
  # Save the plot
  ggsave(file_path, units = "in", dpi = 300, width = 8, height = 6)
}

# List of traits to plot
traits <- colnames(BLUPs_table)[!colnames(BLUPs_table) %in% c("accession_name")]

# Iterate over each trait and apply the function
for (trait in traits) {
  plot_and_save(trait, trait, met_results, trial_interest)
}
```

## Save the BLUEs or raw data across the trials

```{r}
variables <- colnames(BLUPs_table)[!grepl("accession_name", colnames(BLUPs_table))]
for (var in variables) {
  
  cat("\n_______________")
  cat("\nTRIAL:", var, "\n")
  cat("_______________\n")
  
  blue_blup <-
  obj$blues_blups %>%
  select(trial, genotype, trait, BLUEs) %>%
  spread(trait, value = BLUEs) %>%
  select(trial, genotype, any_of(var)) %>%
  group_by(trial, genotype) %>%
  pivot_wider(names_from = trial, values_from = any_of(var)) %>%
  right_join(BLUPs_table %>%
    select(accession_name, any_of(var)), by = c("genotype" = "accession_name")) %>%
  arrange(is.na(across(where(is.numeric))), across(where(is.numeric))) %>%
  mutate(across(where(is.numeric), round, 2))
# remove all NA columns
blue_blup <- blue_blup[, colSums(is.na(blue_blup)) < nrow(blue_blup)]

master_data[[paste0("BLUP_BLUE_", var)]] <- blue_blup
}
```

## Stability analysis

```{r}
for (var in variables) {
  
  cat("\n_______________")
  cat("\nTRIAL:", var, "\n")
  cat("_______________\n")

stab <- met_results$stability %>% 
  filter(trait == var) %>% 
  arrange(superiority) %>% 
        pivot_wider(names_from = "trait", values_from = c('predicted.value')) 

# Change colname
colnames(stab)[5] <- paste('BLUPs', colnames(stab)[5], sep = '_') 
colnames(stab)[c(2, 3, 4)] <- paste(colnames(stab)[c(2, 3, 4)], var, sep = '_') 

master_data[[paste0("stability_", var)]] <- stab
}


ind <- grep("^stability_", names(master_data))


# select elements that satisfy the condition
stab_values <- master_data[ind] %>% 
  reduce(inner_join, by = "genotype") %>% 
  select(!starts_with("BLUPs_")) %>% 
  mutate(across(where(is.numeric), round, 2))

# remove multiple stability sheets
master_data[ind] <- NULL
```

## BLUE and BLUP data together

```{r}
BLUEs_BLUPs <- 
  obj$blues_blups %>%
  select(trait, genotype, trial, BLUEs, seBLUEs) %>%
  filter(trait %in% variables) %>% 
  pivot_wider(names_from = "trait", values_from = c("BLUEs", "seBLUEs")) %>%
  pivot_wider(names_from = trial, values_from = c(
    paste("BLUEs", variables, sep = "_"),
    paste("seBLUEs", variables, sep = "_")
  )) %>%
  left_join(
    met_results$overall_BLUPs %>%
      select(!status) %>%
      rename(
        BLUPs = predicted.value,
        seBLUPs = std.error
      ) %>%
      pivot_wider(names_from = "trait", values_from = c("BLUPs", "seBLUPs")),
    by = "genotype"
  ) %>%
  # arrange(desc(BLUPs_DM_gravity)) %>% 
  arrange(is.na(across(where(is.numeric))), across(where(is.numeric))) %>%
  mutate(across(where(is.numeric), round, 2))
# remove all NA columns
  BLUEs_BLUPs <- BLUEs_BLUPs[, colSums(is.na(BLUEs_BLUPs)) < nrow(BLUEs_BLUPs)]


# put all together stab_values with blues_blups
BLUEs_BLUPs <- 
  BLUEs_BLUPs %>% left_join(stab_values, by = 'genotype')  
  
    
header_sort = vector()
for (i in 1:length(variables)) {
  
  header_sort = c(header_sort, 
                  grep(variables[i], sort(names(BLUEs_BLUPs)), value=TRUE) 
  )
  
}


BLUEs_BLUPs <- BLUEs_BLUPs %>%
  select(genotype, all_of(header_sort), -starts_with("se")) 

# bind with BLUP_BLUE_202338CQEPR_repe
BLUPs_cere <- master_data$BLUP_BLUE_202338CQEPR_repe %>% 
  select(genotype, starts_with("BLUPs"))

colnames(BLUPs_cere)[-1] <- paste(colnames(BLUPs_cere)[-1], "202338CQEPR_repe", sep = "_")

BLUEs_BLUPs <- BLUEs_BLUPs %>% 
  select(-BLUEs_germination_perc_202338CQEPR_repe) %>% left_join(
   BLUPs_cere 
  )

BLUEs_BLUPs %>%
  select(genotype, all_of(header_sort))




master_data[["BLUEs_BLUPs_MET"]] = BLUEs_BLUPs
```

```{r}


# bind with BLUP_BLUE_202338CQEPR_repe
BLUPs_cere <- master_data$BLUP_BLUE_202338CQEPR_repe %>% 
  select(genotype, starts_with("BLUPs"))

colnames(BLUPs_cere)[-1] <- paste(colnames(BLUPs_cere)[-1], "202338CQEPR_repe", sep = "_")

master_data[["BLUPs_cere"]] = BLUPs_cere
```


## Genotypic correlation

```{r}

geno_cor <- gg_cor(
  colours = c("red", "white", "blue"),
  data = BLUPs_cere, # remove carotenoids
  label_size = 3
) + 
  theme(
  axis.text.y = element_text(size = 14),
  axis.text.x = element_text(size = 14))


ggsave(paste("images\\geno_corr", trial_interest, ".png", sep = "_"),
       units = "in", dpi = 300, width = 14, height = 8)
```

## Save the master data results

```{r}
folder_output <- here::here("output//")
meta_file_name <- paste0(folder_output, paste("2023", trial_interest, "master_results", Sys.Date(), ".xlsx", sep = "_"))

write.xlsx(master_data, file = meta_file_name)

# It is neccessary change some elements names from master_file

# grep("^BLUP_BLUE_root_number_commercial", names(master_data))
# names(master_data)[17] <- "BLUP_BLUE_root_commercial"
# 
# grep("^BLUP_BLUE_root_weight_commercial", names(master_data))
# names(master_data)[36] <- "BLUP_BLUE_root_wt_commercial"
# 
# write.xlsx(master_data, file = meta_file_name)
```

## Index selection

```{r}
list_file <- list.files(folder_output)
sel_file <- list_file[str_detect(list_file, "_master_results_") &
  str_detect(list_file, trial_interest)]

sel_file

sel_file[1]
blupDF_kp <- read_excel(
  paste(folder_output,
    sel_file[1],
    sep = ""
  ),
  sheet = paste0("BLUPs_", "cere")
)
```

## Selection index

```{r}
# genotypes <- my_dat %>% 
#   filter(trial_name %in% c("202325LAEAR_saha", 
#                            "202323LAEAR_repe", 
#                            "202324LAEAR_stom")) %>% 
#   distinct(accession_name) %>% pull()

colnames(blupDF_kp)[-1] <- str_remove(colnames(blupDF_kp)[-1], "_202338CQEPR_repe")

index_traits <- c("BLUPs_DM_gravity", "BLUPs_plant_type", "BLUPs_yield_ha", "BLUPs_germination_perc")

index_dat <- blupDF_kp %>%
  select("genotype", all_of(index_traits)) %>% 
  drop_na()

# %>% filter(accession_name %in% all_of(genotypes))
```

## Selection index function

```{r}
# multi-trait -------------------------------------------------------------
library(explor)
library(FactoMineR)
library(factoextra)
library(cowplot)
library(ggpubr)
pca_index <- function(data, id, variables = NULL, percentage = 0.20, b) {
  # The data set to be analyzed. It should be in the form of a data frame.
  data <- as.data.frame(data)
  rownames(data) <- data[, id]
  if (is.null(variables)) variables <- names(data)[names(data) != id]
  data <- data[, variables]
  index <- selIndex(Y = as.matrix(data), b = b, scale = T)
  index <- c(index)
  data$index <- index
  data <- data %>% arrange(desc(index))
  data$selected <- NA
  data$selected[1:(round(percentage * nrow(data)))] <- TRUE
  data$selected <- ifelse(is.na(data$selected), FALSE, data$selected)
  res.pca <- PCA(data, graph = T, scale.unit = T, quali.sup = ncol(data))
  p1 <- fviz_pca_var(res.pca, col.var = "black", repel = T) +
  theme_xiaofei()
  p2 <- fviz_pca_ind(res.pca,
    label = "none", habillage = data$selected,
    palette = c("#00AFBB", "#FC4E07"), addEllipses = T
  ) +
  theme_xiaofei()
  # labs(title =  paste("Selection:",
  #                     paste0(percentage*100,"%")),
  #      subtitle = paste("Weights:", "(", paste0(b, collapse = ', '),")"  ))
  final <- ggdraw() +
  draw_plot(p1, x = 0, y = 0, width = .5, height = 1) +
  draw_plot(p2, x = .5, y = 0.2, width = .5, height = 0.6) +
  draw_plot_label(
    label = c("A", "B"), size = 15,
    x = c(0, 0.5), y = c(1, 1)
  )
  
  # final <- ggarrange(p1, p2, legend = "bottom", common.legend = T)
  # final <- annotate_figure(final,
  #   top = text_grob(paste(
  #     "Selection:",
  #     paste0(percentage * 100, "%"), "\n",
  #     paste("Weights:", "(", paste0(b, collapse = ", "), ")", "\n")
  #   ), color = "black", face = "bold", size = 14)
  # )
  selection <- data %>% filter(selected == T)
  return(list(res.pca = res.pca, final = final, results = data, selection = selection))
}
selIndex <- function (Y, b, scale = FALSE) 
{
    if (scale) {
        return(scale(Y) %*% b)
    }
    return(Y %*% b)
}
```

## Index selection

```{r}

res.pca <- pca_index(data = index_dat, id = "genotype", b = c(10, -5, 10, 15), percentage = 0.15)
res.pca_final <- res.pca$final
res.pca_final
ggsave(paste("images/selection",  trial_interest, ".png"), plot = res.pca_final, units = "in", dpi = 300, width = 10, height = 10)

accessions <- res.pca$selection <- res.pca$selection %>% 
  filter(BLUPs_germination_perc >= 50) %>% 
  arrange(desc(BLUPs_DM_gravity)) %>% row.names()

selections <- res.pca$results %>% rownames_to_column(var = "accession_name") %>% 
  mutate(selected = ifelse(accession_name %in% accessions, TRUE, FALSE)) %>% arrange(desc(BLUPs_germination_perc))

selections %>% 
  select(accession_name, index, everything()) %>% 
  write.table("clipboard", sep = "\t", col.names = T, row.names = F)
```

## Add index column to BLUEs_BLUPs_MET

```{r}
BLUPs_cere <- 
  master_data$BLUPs_cere %>% 
  left_join(selections[-c(2:4)], by = c("genotype" = "accession_name")) %>% 
  relocate(index, selected, .before = 2) %>% 
  arrange(is.na(selected))

master_data[["BLUPs_cere"]] = BLUPs_cere
```

## Save the master data results

```{r}
folder_output <- here::here("output//")
meta_file_name <- paste0(folder_output, paste("2023", trial_interest, "master_results", Sys.Date(), ".xlsx", sep = "_"))
write.xlsx(master_data, file = meta_file_name)
```
